Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-05-14 man/ms.Rmd
update

The forests of this species reaches their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005]. For this reason, the populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], having these forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].

@@

As we mentioned previously, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3). This pattern was also observed for other species in their rear edge [@SanchezSalguero2012; @Camarero2015]. But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6). In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6). Surprisingly this result seems contradict our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge after extreme drought events.

@@

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. During 2005 and 2012 drought events EVI was reduced to 90.25 % (Tabñe 1), whereas BAI was reduced to 70.83 % (Table S2). This suggests a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate. For instance, during 2005 drought event, the BAI were reduced to 44.5 % respect to that of the preceding period in northern site (SJ), whereas the EVI was reduced to 81.9 % for the same site during the same drought event (Tables S1 and S2). These findings are in accordance with previously works which showed that tree-growth are more sensitive metrics of forest resilience than net primary productivty [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018]. Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink thant by source limitations [@Gazol2018]. Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisio2012]. On the other hand, tree-ring width data provide an accurate measure of growth responses to droughts [@Gazol2018].

@@

Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3). Las poblaciones del norte, que presentan condiciones mas secas (Table 1), se vieron mas afectadas por los eventos de sequía. For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2). In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 8), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].

@@

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location [@PerezLuque2011tfm], showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our results are consistent with previous findings that pointed out tree growth and tree responses to drought are site-dependent (e.g. soil features, tree competence, etc.) [e.g. @Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017c].

When we explored the chracteristics of the southern sites, we firstly observed a difference in elevation. @Bhuyan2017, in a recent work across the European continent, found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. In fact, the differences in available water of soils could explain our results. @CoboDiaz2017, in a work analyzing the variation of the microbiota along an elevational gradient within this location, found higher values of available water of the forest soils located on high elevation compared with those located at low elevations. In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply. In most of the high-mountain watersheds of Sierra Nevada, there is a system of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014; @MartinCivantos2016]. These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017]. The network of traditional ditches present in one of our sampling site (Cáñar) was recentlty described, with detailed information about water use [@RuizRuiz2017] and its hydrological functioning [@MartinMontanes2015]. There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017]. It brings water from snowmelt and from a nearby stream, soaking the *Q. pyrenaica* forests. This traditional system could supply an extra of water that could be used by the trees located downstream of these channels. This extra of water is particularly important for trees of this dry-rear edge and could explain the higher values of BAI for trees located in CA-High site.

@@

Our study sites have been subjected to an intense exploitation of forest resources. Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, recorded the quantity of trees, dividing them into three categories: *new*, *growing* and *old* [@Wing2015]. For Cáñar site more than two millions of trees were reported, most of them *news*, and no *old* trees were counted, suggesting recent wood fellings. Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@GomezCruz1992]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; GeaIzquierdo2015; @Dorado2017].

@@

On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019]. Forest resources, like firewood, charcoal, acorns, have been continuously exploited in this site through history [@ValbuenaCarabana2014, @JimenezSerrano2004]. In Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century. From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment [@MesaTorres2009; @Bonet2014_conama].

@@

#### Disturbance
Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).

## historia forestal de ambos sitios 
Incluimos lo que conocemos de la historia forestal de los sitios?? Me explico, tenemos datos de manejo y uso antrópico de las dos zonas que proceden de varias fuentes. En resumen, mas o menos, tenemos:

* Minería: datos de minería, que afectan sobre todo a la población de SJ (los tenemos localizados temporalmente)
* Actuaciones forestales: 
* Incendios, Carboneo, Ganadería, etc --> Tenemos información menos estructurada sobre estos ámbitos, procedentes de varias fuentes, que tienen una incertidumbre mayor espacial y temporalmente. 
")

```

Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution. Para  *Q. pyrenaica* se ha visto que moisture availability es el factor limitante que determina el crecimiento en las poblaciones situadas en su rear edge [@GeaIzquierdo2014], pero para otras especies (*i.e.* *Pinus nigra*, *P. sylvestris*) la temperatura tiene mas peso que la disponibilidad de agua  [@Herrero2013; @Matias2017]. Pero es importante considerar además la historia de manejo que han tenido esos bosques a la hora de la forest management [@Penuelas2017; @DoblasMiranda2017], ya que, como observamos en nuestros resultados puede condicionar el crecimiento y la resiliencia de las especies.

@@

```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
* Vicente-Serrano: 
    * We found that some forests from cold and humid areas respond to shorter drought time-scales than forests from dry areas, which usually respond to longer time-scales
    * Growth and responses to drought are modulated by site conditions such as soil type, specific functional traits and the intensity of competition among neighbouring trees (Orwig & Abrams, 1997; McDowell et al., 2008; Linares et al., 2010; Pasho et al., 2011).
    * Our findings provide evidence that the patterns of growth response to drought do not follow a general geographical structure and that these patterns are driven by the biogeographical, topographic and climatic conditions of each site, showing that forests located in different continents have the same pattern of response to drought time-scales.

>>>>

The forests of this species reaches one of their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada mountain is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]. In this mountain region, *Q.* *pyrenaica* woodlands represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005]. They are considered relictic forests [@Melendo2000; @Vivero2000], and the richest vascular-plant forest formation of Sierra Nevada [@PerezLuqueetal2014P], sheltering several endemic and endangered plant species [@Blanca1998; @Lorite2009]. These forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].

@@

As we mentioned previously, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3), matching similar declining patterns observed for other species in their rear edge [@SanchezSalguero2012; @Camarero2015]. But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6). In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6). Surprisingly this result seems contradict our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge after extreme drought events.

@@

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. During the last drought events EVI was reduced to 90.25 % (Table 1), whereas BAI was reduced to 70.83 % (Table S2), suggesting a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate. For our driest site (SJ, the northern one), BAI were reduced to 44.5 % respect to that of the preceding period during 2005 drought event, whereas the EVI was reduced to 81.9 % (Tables S1 and S2). These findings are in accordance with previously works which showed that tree-growth are more sensitive metrics of forest resilience than net primary productivty [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018]. Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018]. Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisio2012]. On the other hand, tree-ring width data provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].

@@

Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3). The former, which have drier conditions (Table 1), were more impacted by the drought events than the latter. For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2). In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 8), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].

@@

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location [@PerezLuque2011tfm], showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (e.g. soil features, tree competence, etc.) [e.g. @Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017c].

When we explored the chracteristics of the southern sites, we firstly observed a difference in elevation. @Bhuyan2017, in a recent work across the European continent, found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. Soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017]. 
In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply. In most of the high-mountain watersheds of Sierra Nevada, there are systems of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014; @MartinCivantos2016]. These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017]. Several works have recently described the network of traditional ditches for our southern sites (Cáñar), with detailed information about water use [@RuizRuiz2017] and its hydrological functioning [@MartinMontanes2015]. There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017]. It brings water from snowmelt and from a nearby stream, soaking the *Q. pyrenaica* forests. This traditional system could supply an extra of water that could be used by the trees located downstream of these channels. This extra of water is particularly important for trees of this dry-rear edge and could explain the higher values of BAI for trees located in CA-High site.

In addition to the lower water stress, it is likely that the sites at higher elevation had a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources.

@@

Our study sites have been subjected to an intense exploitation of forest resources. Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, recorded the quantity of trees, dividing them into three categories: *new*, *growing* and *old* [@Wing2015]. For Cáñar site more than two millions of trees were reported, most of them *news*, and no *old* trees were counted, suggesting recent wood fellings. Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@GomezCruz1992]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; @GeaIzquierdo2015; @Dorado2017].

@@

On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019]. Forest resources, like firewood, charcoal, acorns, have been continuously exploited in this site through history [@ValbuenaCarabana2014; @JimenezSerrano2004]. In Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century. From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment [@MesaTorres2009; @Bonet2014_conama].

@@

Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution [*e.g.* @GeaIzquierdo2014; @Herrero2013; @Matias2017], but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Penuelas2017; @DoblasMiranda2017; @Navarro2013], since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species.

@@

. Para  *Q. pyrenaica* se ha visto que moisture availability es el factor limitante que determina el crecimiento en las poblaciones situadas en su rear edge [@GeaIzquierdo2014], pero para otras especies (*i.e.* *Pinus nigra*, *P. sylvestris*) la temperatura tiene mas peso que la disponibilidad de agua  [@Herrero2013; @Matias2017]. Pero es importante considerar además la historia de manejo que han tenido esos bosques a la hora de la forest management [@Penuelas2017; @DoblasMiranda2017], ya que, como observamos en nuestros resultados puede condicionar el crecimiento y la resiliencia de las especies.

* Vicente-Serrano: 
    * We found that some forests from cold and humid areas respond to shorter drought time-scales than forests from dry areas, which usually respond to longer time-scales
    * Growth and responses to drought are modulated by site conditions such as soil type, specific functional traits and the intensity of competition among neighbouring trees (Orwig & Abrams, 1997; McDowell et al., 2008; Linares et al., 2010; Pasho et al., 2011).
    * Our findings provide evidence that the patterns of growth response to drought do not follow a general geographical structure and that these patterns are driven by the biogeographical, topographic and climatic conditions of each site, showing that forests located in different continents have the same pattern of response to drought time-scales.
